DOpElib: A Goal Oriented Software Library for Computing PDEs and Optimization Problems

### 2012-10-04 Paper_May_2012/main.tex
Modifikationen am Artikel: Kommentare und Ideen von der Diskussion mit Christian eingefuegt.
Paper: Kapitel 2 mit Beschreibung der main features begonnen. Teil fuer stationaere Probleme fast fertig. Kapitel 3: FSI Example fertig gestellt. D.h. kurze Beschreibung und Darstellung der wesentlichen Ergebnisse.

%
%%% User-requested packages placed after this line %%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amsfonts,amssymb,amsmath}
\usepackage{booktabs,dcolumn}
\usepackage[T1]{fontenc}
\usepackage{mathtools}
\usepackage{enumerate,graphicx}
\usepackage{subfig}
\usepackage{multirow}
% \usepackage[center]{subfigure}
\numberwithin{equation}{section}
\usepackage{braket}                             %Provides \Set for typing sets
\usepackage{float}
\usepackage{subfig}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Detailed Description of the Features}
\label{detailed_description}
\todo{Aus der Sicht des Users anhand eines konkreten Beispiels beschreiben}
\todo{Einfaches Beispiel nehmen}
\todo{erwaehnen, dass wir alles immer nicht-linear betrachten}
\todo{Aussehen eines typischen PDE und Opt loop und deren Gemeinsamkeiten
erklaeren. Vielleicht mir Pseudocode}
This library is designed to allow easy implementation and numerical solutions 
of problems involving partial differential equations (PDEs). The easiest case 
is that of a PDE in weak form to find some $u$
\[
a(u)(\phi) = 0 \quad \forall \phi \in V,
\]
with some appropriate space $V$.
More complex cases involve optimization problems given in the form (OPT)
\begin{align*}
\min\;&J(q,u) \\
  &\text{s.t.}\; a(q,u)(\phi) = 0 \quad \forall \phi\in V,\\
  &a \le q \le b,\\
  &g(q,u) \le 0,  
\end{align*}
where $u$ is a FE-function and $q$ can either be a FE-function or some 
fixed number of parameters, $a$ and $b$ are constraint bounds for the control $q$,
and $g(\cdot)$ is some state constraint.

\subsection{Problem description}
In order to allow our algorithms the automatic assembly of all required 
data we need to have some container which contains the complete problem 
description in a common data format. For this we have the following 
classes in \texttt{DOpEsrc/container}
\begin{itemize}
  \item \texttt{pdeproblemcontainer.h} Is used to describe  stationary PDE problems.
  \item \texttt{instatpdeproblemcontainer.h} {\bf TODO is still missing but will be used for nonstationary problems.} This will be implemented once we have nonstationary optimization problems running to avoid error duplication in the coding process.
  \item \texttt{optproblem.h} Is used to describe  OPT problems governed by 
    stationary PDEs. 
  \item \texttt{instatoptproblemcontainer.h} Is used to describe  OPT problems
    governed by nonstationary PDEs. The only difference to the stationary case
    is that we need to specify a time-stepping method.  
  \item \texttt{interfaces/functionalinterface.h} This gives an interface 
    for the functional $J(\cdot)$ and any other functional you may want to evaluate.
    In general this can be used as a base class to write your own functionals 
    in examples. We note that we only need to write the integrands on 
    elements or faces the loop over elements will be taken care of in the integrator.
    Specifically, derivatives are written therein, too.
\end{itemize}
In order to fill these containers there are two things to be done,
first we need to actually write some data, for instance,
the semilinear form $a(\cdot)(\cdot)$, a target functional $J(\cdot)$, etc.,
which describe the problem. Then we have to select some numerical 
algorithm components like finite elements, linear solvers $\ldots$.
The latter ones should be written such that when exchanging these components
none of the problem descriptions should require changes. 
Note that it still may be necessary to write some additional descriptions, 
e.g., if you solve the PDE with a fix point iteration you don't need derivatives
but if you want to use Newton's method, derivatives are needed.

We will start by discussing the problem description components implemented so far

\subsection{Numeric components}
These are the components from which a user needs to select some in order to actually 
solve the given problem. They will not require any rewriting, but sometimes it is 
advisable to write other than the default parameter into the param file for the 
solution.

\subsubsection{Space-time handler}
First we need to select a method how to handle all dofs in space and time.
\begin{itemize}
\item \texttt{basic/spacetimehandler\underline{ }base.h} This class is used to define 
  an interface to the dimension independent functionality of all space time dof handlers.
  {\bf TODO: Beispiele geben}
\item \texttt{basic/statespacetimehandler.h} Another intermediate interface class which adds 
  the dimension dependent functionality if only the variable $u$ is considered, i.e., a 
  PDE problem.
\item \texttt{basic/spacetimehandler.h } Same as above but with both $q$ and $u$, i.e., for
  OPT problems.
\item \texttt{basic/mol\underline{ }statespacetimehandler.h} Implementation of a method of 
  line space time dof handler for PDE problems. It has only one spatial 
  dofhandler that is used for all time intervals.
\item \texttt{basic/mol\underline{ }spacetimehandler.h} Same as above for OPT problems.
  A separate spatial dof handler for each of the variables $q$ and $u$ is maintained 
  but only one triangulation.
\item \texttt{basic/mol\underline{ }multimesh\underline{ }spacetimehandler.h}
  Same as above, but now in addition the triangulations for $q$ and $u$ can be refined
  separately from one common initial coarse triangulation. Note that this will
  in addition require the use of the multimesh version for integrator and 
  face- as well as celldatacontainer.
\end{itemize}
Note that we use these for stationary problems as well, but then you don't have to specify
any time information.

\subsubsection{Container classes}
Second you will need to specify some container classes to be used to 
pass data between objects. At present you don't have much choice, but you may wish 
to reimplement some of these if you need data that is not currently included in 
the containers.
\begin{itemize}
\item \texttt{container/celldatacontainer.h} This object is used to pass data 
  given on the current element (cell) of the mesh to the functions in PDE, functional, 
  $\ldots$. 
\item \texttt{container/facedatacontainer.h} This object is used to pass data 
  given on the current face of the mesh to the functions in PDE, functional, 
  $\ldots$. 
\item \texttt{container/multimesh\underline{ }celldatacontainer.h} This is the same as the 
  celldatacontainer, but it
  is capable to handle data defined on an alternative triangulation.
\item \texttt{container/multimesh\underline{ }facedatacontainer.h} This is the same as the
  facedatacontainer, but it
  is capable to handle data defined on an alternative triangulation.
\item \texttt{container/integratordatacontainer.h} This contains some data that 
  should be passed to the integrator like quadrature formulas and the above cell and 
  face data container.
\end{itemize}

\subsubsection{Time stepping schemes}
\todo{Kurz halten}
Third, at least for nonstationary PDEs we need to select a time stepping scheme
the file names of which are mostly self explanatory:
\begin{itemize}
\item \texttt{include/forward\underline{ }euler\underline{ }problem.h}
\item \texttt{include/shifted\underline{ }crank\underline{ }nicolson\underline{ }problem.h}
\item \texttt{include/backward\underline{ }euler\underline{ }problem.h}
\item \texttt{include/fractional\underline{ }step\underline{ }theta\underline{ }problem.h} Note that the use of this scheme requires a special Newton solver, which is, however, already
implemented for the convenience of the user!
\item \texttt{include/crank\underline{ }nicolson\underline{ }problem.h}
\end{itemize}

\subsubsection{Integrator routines}
Finally, we need to select a way how to integrate and solve linear and nonlinear equations
\begin{itemize}
\item \texttt{templates/integrator.h} This class computes integrals over a given 
  triangulation (including its faces).
\item \texttt{templates/integrator\underline{ }multimesh.h} The same as above but it is 
  possible that some of the FE functions are defined on an other triangulation 
  as long as the have a common coarse triangulation.
\item \texttt{templates/integratormixeddims.h} This is used to compute integral which 
  are given in another (larger) dimension than the current variable. (This is exclusively
  used if the control variable is given by some parameters. Which means \texttt{dopedim == 0}). 
\end{itemize}

\subsubsection{Nonlinear solvers}
\begin{itemize}
\item \texttt{templates/newtonsolver.h} This solves some nonlinear equation using a 
  line-search Newton method.
\item \texttt{templates/newtonsolvermixeddims.h} The same but in the case when there is 
  another variable in a (larger) dimension is involved. See 
  \texttt{integratormixeddims.h}.
\item \texttt{templates/instat\underline{ }step\underline{ }newtonsolver.h} This is a 
  Newton method as above to invert the next time-step. It differs from the plain vanilla
  version in that it computes certain data from the previous time step only once 
  and not in every Newton iteration.
\item \texttt{templates/fractional\underline{ }step\underline{ }theta\underline{ }step\underline{ }newtonsolver.h} This is the Newton solver for the time step in a 
  fractional-step-theta scheme. It combines the computation of all three sub steps.
\end{itemize}

\subsubsection{Linear solvers}
To solve the linear equations within Newton's method, we use 
standard solvers from iterative type 
such as the method of the conjugate gradients (CG method), 
GMRES, or direct solvers such as \texttt{UMFPACK}.
In addition, we DOpE provides  a wrapper class 
\texttt{templates/voidlinearsolver.h} for certain cases when we 
  know that the matrix to be inverted is the identity. It simply copies the rhs to the
  lhs. This is only needed for compatibility reasons some other components.

%\begin{itemize}
%\item \texttt{templates/cglinearsolver.h} This is a wrapper for the cg solver implemented in 
%  \texttt{deal.II}. The solver will build and store the stiffness matrix for the PDE.
%\item \texttt{templates/gmreslinearsolver.h} This is a wrapper for the gmres solver 
%  implemented in \texttt{deal.II}. The solver will build and store the stiffness matrix 
%  for the PDE.
%\item \texttt{templates/directlinearsolver.h} This is a wrapper for the direct solver 
%  implemented in \texttt{deal.II} using \texttt{UMFPACK}. 
%  The solver will build and store the stiffness matrix for the PDE.
%\item \texttt{templates/voidlinearsolver.h} This is a wrapper for certain cases when we 
%  know that the matrix to be inverted is the identity. It simply copies the rhs to the
%  lhs. This is only needed for compatibility reasons some other components.
%\end{itemize}
%

%\subsection{Problem specific classes}
%The following classes are used to describe the problem and will usually require 
%some implementation.
%
%\begin{itemize}
%  \item \texttt{basic/constraints.h} This is used by the spacetimehandlers to 
%    compute the number of constraints from the control and state vectors. 
%    It must not be reimplemented by the user, but needs to be properly 
%    initialized if OPT is used with box control constraints or $g(q,u) \le 0$.
%
%  \item \texttt{interfaces/constraintinterface.h} This gives an interface for both 
%    the control box constraints as well as the general constraint $g \le 0$. This 
%    needs to be specified if constraints are to be used. If they are not needed 
%    a default class \texttt{problemdata/noconstraints.h} can be used. We note that we only 
%    need to write the integrands on 
%    elements or faces the loop over elements will be taken care of in the integrator.
%  \item \texttt{interfaces/pdeinterface.h} This defines an interface for the 
%    partial differential equation $a(q,u)(\phi) = 0$. This needs to be written
%    by the user. We note that we only need to write the integrands on 
%    elements or faces the loop over elements will be taken care of in the integrator.
%    Specifically, derivatives are written therein, too.
%  \item \texttt{interfaces/dirichletdatainterface.h} This gives an interface to the 
%    Dirichlet data for a problem. If the Dirichlet data are simply a function 
%    (and do not depend on the control $q$) one can use the default class\\
%    \texttt{problemdata/simpledirichletdata.h}.
%\end{itemize}
%

\subsection{Reduced problems (Solve the PDE)}
At times it is nice to remove the PDE constraint in (OPT). 
This is handled by so called reduced 
problems (for algorithmic aspects we refer the reader to 
\cite{BeMeVe06}). 
This means that the reduced problem implicitly solves the PDE whenever required
and eliminates the variable $u$ from the problem.
\begin{itemize}
\item \texttt{reducedproblems/statpdeproblem.h} This is used to remove the variable $u$ in 
  a stationary PDE problem. This means that call the method \\
  \texttt{StatPDEProblem::ComputeReducedFunctionals} will evaluate the functionals 
  defined in the problem description, i.e., in \texttt{PDEProblemContainer}, in the 
  solution of the given PDE.
\item \texttt{reducedproblems/statreducedproblem.h} This eliminates $u$ from the OPT
  problem with a stationary PDE.
\item \texttt{reducedproblems/instatreducedproblem.h} The same as above but for a
  nonstationary PDE. {\bf FIXME there is something wrong in this file see FIXME 
    comment in the source.}
\item \texttt{reducedproblems/voidreducedproblem.h} A wrapper file that eliminates $u$ 
  if it is not present anyways. This is used so that we can use the same routines to 
  solve problems that have no PDE constraint.
\end{itemize}

\subsection{Optimization algorithms}
Now, in order to solve optimization algorithms we need to define some algorithms.
At present we offer a selection of algorithms that solve the reduced optimization 
problem where the PDE constraint has been eliminated as explained in the previous section.
\begin{itemize}
\item \texttt{opt\underline{ }algorithms/reducedalgorithm.h} An interface for all 
  optimization problems in the reduced formulation. It offers some test functionality
  to assert that the derivatives of the problem are computed correctly.
\item \texttt{opt\underline{ }algorithms/reducednewtonalgorithm.h}
  A line-search Newton algorithm using a cg method to invert the reduced hessian. 
  Implementation ignores any additional constraints.
\item \texttt{opt\underline{ }algorithms/reducedtrustregionnewton.h}
  A trust region Newton algorithm using a cg method to invert the reduced hessian.
  Implementation ignores any additional constraints.
\item \texttt{opt\underline{ }algorithms/reduced\underline{ }snopt\underline{ }algorithm.h}
  An algorithm to solve reduced optimization problems with additional control constraints.
  ((reduced) state constraints are not yet implemented.)
\item \texttt{opt\underline{ }algorithms/reducednewtonalgorithmwithinverse.h}
  Line-search Newton algorithm that assumes there exists a method in the reduced problem
  that can invert the reduced hessian. (This usually makes sense only if there is no 
  PDE constraint.)
\item \texttt{opt\underline{ }algorithms/generalized\underline{ }mma\underline{ }algorithm.h}
  An implementation of the MMA-Algorithm for structural optimization using an augmented
  Lagrangian formulation for the subproblems. The subproblem is implemented using the 
  special purpose file
  \texttt{include/augmentedlagrangianproblem.h}.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Interfaces to other software packages}
Since DOpE uses as basis the software library deal.II, all its interfaces
to other packages can be accessed to. For instance, such libararies 
are trilinos for with an algebraic multigrid method, MPI for a parallel
solution of the problem.

In addition DOpE itself has an interface to SNOPT.

\section{Documentation, Code Development, Webpage}
\label{documentation}
At the present step, the DOpElib project comes with 
an detailed documentation of all features and examples 
in pdf format and an detailed programming code 
documentation.

\todo{License? How to access to DOpE?}
\todo{Name of the website}

In addition, the DOpElib test suite provides some regression tests. 
They are run to compare 
the output to previous outputs. This is useful (necessary) after 
changing programming code anywhere in the library. If a test
succeeds, everything is fine in the library. If not, you should not
check in your code into DOpE. Please make sure what is going wrong and WHY!
Every command is computed via a Makefile.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Applications}
\label{applications}
We present some numerical tests that have been already computed 
with DOpElib. These tests demonstrate the performance 
of DOpElib by covering different computational aspects. 
\begin{itemize}
\item The first numerical test shows various time-stepping schemes
and the numerical solution of nonlinear coupled partial differential
equations, formulated in a monolithic solution algorithm. 
Specifically, the equations are solved by the method of lines, i.e,
first spatial discretization and then temporal  discretization 
with a Galerkin finite element scheme. The nonlinaer problem
is solved by Newton's method. The Jacobian is derived 
by analytical expressions of the derivatives. 
In DOpElib this example is implemented in 
\texttt{Example/PDE/InstatPDE/Example2}
\item The second example presents an
optimization problem and its numerical solution.
\item The main purpose of the third example is the application 
of refinement techniques for local mesh refinement for 
two and three dimensional problems. We consider the numerical
solution of fluid flows governed by the Navier-Stokes
equations. In contrast to the previous tests, we 
use finite elements on locally adapted meshes 
with goal-oriented mesh refinement with the help
of the dual-weighted residual method. Using 
this method of local mesh refinement, the error estimator
is derived with respect to some goal functional and 
provides optimal meshes to the accurate measurement 
of the considered functional value. Consequently,
this is an effiecient method to reduce the computational 
cost. Most other 
software libraries does not provide goal-oriented 
mesh refinement.  
\item Here, we show a time-dependent optimization problem or 
FSI optimization. 
\end{itemize}

\subsection{Nonstationary Fluid-Structure Interaction}
In this example, the three proposed mesh motion models are applied 
to an unsteady fluid-structure interaction problem. We consider 
the numerical benchmark test FSI 2, which was proposed in \cite{HrTu06b}. 
The configuration is 
sketched in Figure \ref{configuration_csm_and_fsi_2D}.
New results can be found in \cite{BuSc06, TuHrMaRaWoAc10, DeHaeAnnBrVie10}. 
The Fractional-Step-$\theta$ scheme was used for time discretization with
different time step sizes $k$.

>>>>

%
%%% User-requested packages placed after this line %%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amsfonts,amssymb,amsmath}
\usepackage{booktabs,dcolumn}
\usepackage[T1]{fontenc}
\usepackage{mathtools}
\usepackage{enumerate,graphicx}
\usepackage{subfig}
\usepackage{multirow}
\usepackage{listings}

% \usepackage[center]{subfigure}
\numberwithin{equation}{section}
\usepackage{braket}                             %Provides \Set for typing sets
\usepackage{float}
\usepackage{subfig}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Detailed Description of the Main Features}
\label{detailed_description}
\todo{Aus der Sicht des Users anhand eines konkreten Beispiels beschreiben}
\todo{Einfaches Beispiel nehmen}
\todo{erwaehnen, dass wir alles immer nicht-linear betrachten}
\todo{Aussehen eines typischen PDE und Opt loop und deren Gemeinsamkeiten
erklaeren. Vielleicht mir Pseudocode}
This library is designed to allow easy implementation and numerical solutions 
of problems involving partial differential equations (PDEs). 
\subsection{Solving of PDEs}
\subsubsection{Stationary problems}
The easiest case 
is that of a PDE in weak form to find some $u$
\[
a(u)(\phi) = 0 \quad \forall \phi \in V,
\]
with some appropriate space $V$. As example, we take 
the Laplace problem in weak form:
\[
a(u)(\phi) = (\nabla u, \nabla \phi) \quad \forall \phi \in V.
\]
To solve the problem, we need a configuration (a mesh), eventually 
a diffusion parameter, a finite element, a quadrature rule,
a linear solver (here the CG solver), and some output such as
a graphical solution and eventually some measurement functional.
Each of those routines are written in the \texttt{main.cc} file. 
We start by including necessary files both the DOpElib library and
the deal.II software:
\begin{lstlisting}
#include "pdeproblemcontainer.h" etc. ...
\end{lstlisting}
In the \texttt{main} function, the following steps need to be done. First, 
the parameter reader for runtime parameters is initialized:
\begin{lstlisting}
  ParameterReader pr;
  SSolver::declare_params(pr);
  DOpEOutputHandler<VECTOR>::declare_params(pr);
  pr.read_parameters(paramfile);
\end{lstlisting}
A linear finite element in two dimensions and two components is 
initialized:
\begin{lstlisting}
  FESystem<2> state_fe(FE_Q<2>(1), 2);
\end{lstlisting}
and a quadrature rule of order three is used to approximate the 
integrals:
\begin{lstlisting}
  QGauss<2> quadrature_formula(3);
\end{lstlisting}
Then, the \texttt{LocalPDE} object comes into play. This is an quite 
important feature of DOpElib because it allows for easy implementation 
of the variational formulation. Specific explication is given below because
the implementation is seperated in \texttt{localpde.h}.
Afterwards, a \texttt{LocalPointFunctional} is declared. As for the PDE, the 
implementation is done in another file, namely \texttt{localfunctional.h}.

Generically, each problem in DOpElib is considered to be time-dependent and nonlinear because
these are the 'real' world problems. Therefore, (even for the linear and stationary Laplace equation),
we have to declare a (pseudo) time:
\begin{lstlisting}
  std::vector<double> times(1, 0.);
\end{lstlisting}
Then, the spatial mesh (here the unit square) is created and three-times globally-refined:
\begin{lstlisting}
  GridGenerator::hyper_cube(triangulation, 0, 1);
  triangulation.refine_global(3);
\end{lstlisting}
A major component which organizes the whole solution process (independent from 
the problem!) is the 
\begin{lstlisting}
MethodOfLines_StateSpaceTimeHandler<FE, DOFHANDLER, SPARSITYPATTERN, VECTOR, 2> DOFH(
      triangulation, state_fe);
\end{lstlisting}
Then, the problem (here the Laplace PDE) is given to the \texttt{DOFH} via
\begin{lstlisting}
  OP P(LPDE, DOFH);
\end{lstlisting}
It remains to declare boundary conditions for the problem under consideration. 
Let's simply take zero Dirichlet conditions on the boundary:
\begin{lstlisting}
  std::vector<bool> comp_mask(2);

  comp_mask[0] = true;
  comp_mask[1] = true;

  DOpEWrapper::ZeroFunction<2> zf(2);
  SimpleDirichletData<VECTOR, 2> DD1(zf);

  P.SetDirichletBoundaryColors(0, comp_mask, &DD1);
\end{lstlisting}
Next, the solver for the linear equations needs to be specified:
\begin{lstlisting}
  SSolver solver(&P, "fullmem", pr, idc);
\end{lstlisting}
Finally, the ouput handler is initialized via
\begin{lstlisting}
  P.RegisterOutputHandler(&out);
  solver.RegisterOutputHandler(&out);
\end{lstlisting}
Finally, the programm is started with:
\begin{lstlisting}
      solver.ReInit();
      out.ReInit();
      out.Write(outp, 1, 1, 1);

      solver.ComputeReducedFunctionals();
\end{lstlisting}
After explaining the basic compononents that are necessary for all  
(linear and stationary) partial differential equations, we shall give 
an overview to the implementation of the specific form of the PDE. 
This step is done in the \texttt{localpde.h} file. As already explained,
each PDE is treatet in a nonlinear manner such that for solving 
the Laplace problem the following nonlinear step is necessary to recapitulate:
Find $\delta u\in V$:
\[
(\nabla \delta u, \nabla \phi) = -(\nabla u, \nabla \phi) 
\]
and $u^{n+1} = u^n + \delta u$ for $n=1,2,3,ldots$.
Therefore, the implementation reads for the right hand side:
\begin{lstlisting}
    void CellEquation(const CellDataContainer<dealii::DoFHandler<dealdim>, VECTOR, dealdim>& cdc,
                      dealii::Vector<double> &local_cell_vector,
                      double scale, double)
    {
      // implement (\nabla u, \nabla \phi) 
    }
\end{lstlisting}
Specifically,
\begin{lstlisting}
      for (unsigned int q_point = 0; q_point < n_q_points; q_point++)
      {
        Tensor<2, 2> ugrads;
        ugrads.clear();
        ugrads[0][0] = _ugrads[q_point][0][0];
        ugrads[0][1] = _ugrads[q_point][0][1];
        ugrads[1][0] = _ugrads[q_point][1][0];
        ugrads[1][1] = _ugrads[q_point][1][1];

	for (unsigned int i = 0; i < n_dofs_per_cell; i++)
        {
	  const Tensor<2, 2> phi_i_grads_u = state_fe_values[displacements].gradient(i, q_point);

          local_cell_vector(i) += scale * scalar_product(ugrads, phi_i_grads_u) 
	    * state_fe_values.JxW(q_point);
        }
      }
\end{lstlisting}
For the left hand side, we need to implement the Jacobian matrix:
\begin{lstlisting}
    void CellMatrix(const CellDataContainer<dealii::DoFHandler<dealdim>, VECTOR, dealdim>& cdc, FullMatrix<double> &local_entry_matrix, double, double)
    {
// implement  (\nabla \delta u, \nabla \phi)
}
\end{lstlisting}
Specifically,
\begin{lstlisting}
      for (unsigned int q_point = 0; q_point < n_q_points; q_point++)
      {
        for (unsigned int k = 0; k < n_dofs_per_cell; k++)
        {
          phi_grads_u[k] = state_fe_values[displacements].gradient(k, q_point);
        }

        for (unsigned int i = 0; i < n_dofs_per_cell; i++)
        {
          for (unsigned int j = 0; j < n_dofs_per_cell; j++)
          {
            local_entry_matrix(i, j) += scalar_product(phi_grads_u[j], phi_grads_u[i]) 
                * state_fe_values.JxW(q_point);
          }
        }
      }
\end{lstlisting}
The cost functional is now implemented in \todo{hier morgen weiter machen}

\subsubsection{Nonstationary problems}
In this section, we focus on the extension to 
the numerical solution of nonstationary processes. 
As example, we consider the heat equation:

...

Because the \texttt{main} file 
is basically the same as for stationary problems
which is one major goal of the software library,
we only focus on the implementation of the PDE in the 
\texttt{localpde.h} file. Here, the stationary part, i.e.,
the diffusion is implemented as before. The only thing we add 
is the time-depdendent part $\partial_t u$:

\begin{lstlisting}
 void CellTimeEquation(const CellDataContainer<dealii::DoFHandler<dealdim>, VECTOR, dealdim>& cdc,
                           dealii::Vector<double> &local_cell_vector __attribute__((unused)),
		       double scale __attribute__((unused)))
    {
// implement \partial_t u in weak form
}
\end{lstlisting}
and it is done in a straight-forward way:
\begin{lstlisting}
 for (unsigned int q_point=0;q_point<n_q_points;q_point++)
	{
	  Tensor<1,2> v;
	  u[0] = _uvalues[q_point](0);
	  u[1] = _uvalues[q_point](1);

	  for (unsigned int i=0;i<n_dofs_per_cell;i++)
	    {
	      const Tensor<1,2> phi_i_u = state_fe_values.value (i, q_point);

	      local_cell_vector(i) +=  scale * (u * phi_i_u) * state_fe_values.JxW(q_point);
	    }
	}
\end{lstlisting}

\subsection{Solving of optimization problems}
More complex cases involve optimization problems given in the form (OPT)
\begin{align*}
\min\;&J(q,u) \\
  &\text{s.t.}\; a(q,u)(\phi) = 0 \quad \forall \phi\in V,\\
  &a \le q \le b,\\
  &g(q,u) \le 0,  
\end{align*}
where $u$ is a FE-function and $q$ can either be a FE-function or some 
fixed number of parameters, $a$ and $b$ are constraint bounds for the control $q$,
and $g(\cdot)$ is some state constraint.

\subsection{Independent features}
In this final section, we shall give a brief summary of features 
that are applicable (in general) to all types of problems. Specifically,
DOpElib provides goal-oriented mesh refinement with the help of the 
Dual-Weighted-Residual (DWR) method \cite{BeRa96}. Here, a certain 
cost functional such as a point evaluation, line intergration or domain
integration is considered. Determining the lowest error between the 
(unknown) exact solution and its numerical approximation for lowest computational
cost is the main goal of the DWR method.

Moreover, DOpElib provides standard error estimators based on residual evaluation and 
a ZZ error estimator, i.e., measureing the smoothness of the discrete first derivatives.
The latter one is accessed via deal.II.

In addition, DOpElib provides multi-mesh procedures 
\todo{kann Winni hier etwas schreiben?}

Concerning temporal integration schemes, the library comes 
with various time stepping schemes (based on finite differences), 
such as the first-order forward Euler and first-order backward Euler scheme. 
To compute fully nonstationary processes, the second-order schemes
Crank-Nicolson (trapozoidal rule), 
shifted Crank-Nicolson (slight shift to the implicit side to maintain
global stability), 
and the Fractional-Step-$\Theta$ scheme, are implemented.

%\subsection{Problem description}
%In order to allow our algorithms the automatic assembly of all required 
%data we need to have some container which contains the complete problem 
%description in a common data format. For this we have the following 
%classes in \texttt{DOpEsrc/container}
%\begin{itemize}
%  \item \texttt{pdeproblemcontainer.h} Is used to describe  stationary PDE problems.
%  \item \texttt{instatpdeproblemcontainer.h} {\bf TODO is still missing but will be used for nonstationary problems.} This will be implemented once we have nonstationary optimization problems running to avoid error duplication in the coding process.
%  \item \texttt{optproblem.h} Is used to describe  OPT problems governed by 
%    stationary PDEs. 
%  \item \texttt{instatoptproblemcontainer.h} Is used to describe  OPT problems
%    governed by nonstationary PDEs. The only difference to the stationary case
%    is that we need to specify a time-stepping method.  
%  \item \texttt{interfaces/functionalinterface.h} This gives an interface 
%    for the functional $J(\cdot)$ and any other functional you may want to evaluate.
%    In general this can be used as a base class to write your own functionals 
%    in examples. We note that we only need to write the integrands on 
%    elements or faces the loop over elements will be taken care of in the integrator.
%    Specifically, derivatives are written therein, too.
%\end{itemize}
%In order to fill these containers there are two things to be done,
%first we need to actually write some data, for instance,
%the semilinear form $a(\cdot)(\cdot)$, a target functional $J(\cdot)$, etc.,
%which describe the problem. Then we have to select some numerical 
%algorithm components like finite elements, linear solvers $\ldots$.
%The latter ones should be written such that when exchanging these components
%none of the problem descriptions should require changes. 
%Note that it still may be necessary to write some additional descriptions, 
%e.g., if you solve the PDE with a fix point iteration you don't need derivatives
%but if you want to use Newton's method, derivatives are needed.
%
%We will start by discussing the problem description components implemented so far
%
%
%\subsection{Numeric components}
%These are the components from which a user needs to select some in order to actually 
%solve the given problem. They will not require any rewriting, but sometimes it is 
%advisable to write other than the default parameter into the param file for the 
%solution.
%
%\subsubsection{Space-time handler}
%First we need to select a method how to handle all dofs in space and time.
%\begin{itemize}
%\item \texttt{basic/spacetimehandler\underline{ }base.h} This class is used to define 
%  an interface to the dimension independent functionality of all space time dof handlers.
%  {\bf TODO: Beispiele geben}
%\item \texttt{basic/statespacetimehandler.h} Another intermediate interface class which adds 
%  the dimension dependent functionality if only the variable $u$ is considered, i.e., a 
%  PDE problem.
%\item \texttt{basic/spacetimehandler.h } Same as above but with both $q$ and $u$, i.e., for
%  OPT problems.
%\item \texttt{basic/mol\underline{ }statespacetimehandler.h} Implementation of a method of 
%  line space time dof handler for PDE problems. It has only one spatial 
%  dofhandler that is used for all time intervals.
%\item \texttt{basic/mol\underline{ }spacetimehandler.h} Same as above for OPT problems.
%  A separate spatial dof handler for each of the variables $q$ and $u$ is maintained 
%  but only one triangulation.
%\item \texttt{basic/mol\underline{ }multimesh\underline{ }spacetimehandler.h}
%  Same as above, but now in addition the triangulations for $q$ and $u$ can be refined
%  separately from one common initial coarse triangulation. Note that this will
%  in addition require the use of the multimesh version for integrator and 
%  face- as well as celldatacontainer.
%\end{itemize}
%Note that we use these for stationary problems as well, but then you don't have to specify
%any time information.
%
%\subsubsection{Container classes}
%Second you will need to specify some container classes to be used to 
%pass data between objects. At present you don't have much choice, but you may wish 
%to reimplement some of these if you need data that is not currently included in 
%the containers.
%\begin{itemize}
%\item \texttt{container/celldatacontainer.h} This object is used to pass data 
%  given on the current element (cell) of the mesh to the functions in PDE, functional, 
%  $\ldots$. 
%\item \texttt{container/facedatacontainer.h} This object is used to pass data 
%  given on the current face of the mesh to the functions in PDE, functional, 
%  $\ldots$. 
%\item \texttt{container/multimesh\underline{ }celldatacontainer.h} This is the same as the 
%  celldatacontainer, but it
%  is capable to handle data defined on an alternative triangulation.
%\item \texttt{container/multimesh\underline{ }facedatacontainer.h} This is the same as the
%  facedatacontainer, but it
%  is capable to handle data defined on an alternative triangulation.
%\item \texttt{container/integratordatacontainer.h} This contains some data that 
%  should be passed to the integrator like quadrature formulas and the above cell and 
%  face data container.
%\end{itemize}
%
%
%\subsubsection{Time stepping schemes}
%\todo{Kurz halten}
%Third, at least for nonstationary PDEs we need to select a time stepping scheme
%the file names of which are mostly self explanatory:
%\begin{itemize}
%\item \texttt{include/forward\underline{ }euler\underline{ }problem.h}
%\item \texttt{include/shifted\underline{ }crank\underline{ }nicolson\underline{ }problem.h}
%\item \texttt{include/backward\underline{ }euler\underline{ }problem.h}
%\item \texttt{include/fractional\underline{ }step\underline{ }theta\underline{ }problem.h} Note that the use of this scheme requires a special Newton solver, which is, however, already
%implemented for the convenience of the user!
%\item \texttt{include/crank\underline{ }nicolson\underline{ }problem.h}
%\end{itemize}
%
%\subsubsection{Integrator routines}
%Finally, we need to select a way how to integrate and solve linear and nonlinear equations
%\begin{itemize}
%\item \texttt{templates/integrator.h} This class computes integrals over a given 
%  triangulation (including its faces).
%\item \texttt{templates/integrator\underline{ }multimesh.h} The same as above but it is 
%  possible that some of the FE functions are defined on an other triangulation 
%  as long as the have a common coarse triangulation.
%\item \texttt{templates/integratormixeddims.h} This is used to compute integral which 
%  are given in another (larger) dimension than the current variable. (This is exclusively
%  used if the control variable is given by some parameters. Which means \texttt{dopedim == 0}). 
%\end{itemize}
%
%\subsubsection{Nonlinear solvers}
%\begin{itemize}
%\item \texttt{templates/newtonsolver.h} This solves some nonlinear equation using a 
%  line-search Newton method.
%\item \texttt{templates/newtonsolvermixeddims.h} The same but in the case when there is 
%  another variable in a (larger) dimension is involved. See 
%  \texttt{integratormixeddims.h}.
%\item \texttt{templates/instat\underline{ }step\underline{ }newtonsolver.h} This is a 
%  Newton method as above to invert the next time-step. It differs from the plain vanilla
%  version in that it computes certain data from the previous time step only once 
%  and not in every Newton iteration.
%\item \texttt{templates/fractional\underline{ }step\underline{ }theta\underline{ }step\underline{ }newtonsolver.h} This is the Newton solver for the time step in a 
%  fractional-step-theta scheme. It combines the computation of all three sub steps.
%\end{itemize}
%
%\subsubsection{Linear solvers}
%To solve the linear equations within Newton's method, we use 
%standard solvers from iterative type 
%such as the method of the conjugate gradients (CG method), 
%GMRES, or direct solvers such as \texttt{UMFPACK}.
%In addition, we DOpE provides  a wrapper class 
%\texttt{templates/voidlinearsolver.h} for certain cases when we 
%  know that the matrix to be inverted is the identity. It simply copies the rhs to the
%  lhs. This is only needed for compatibility reasons some other components.
%
%
%
%
%%\begin{itemize}
%%\item \texttt{templates/cglinearsolver.h} This is a wrapper for the cg solver implemented in 
%%  \texttt{deal.II}. The solver will build and store the stiffness matrix for the PDE.
%%\item \texttt{templates/gmreslinearsolver.h} This is a wrapper for the gmres solver 
%%  implemented in \texttt{deal.II}. The solver will build and store the stiffness matrix 
%%  for the PDE.
%%\item \texttt{templates/directlinearsolver.h} This is a wrapper for the direct solver 
